---
title: "Measuring multi-year changes in the Symbiodiniaceae algae in Caribbean corals on coral-depleted reefs"
author: "R. Cunning"
date: "08/27/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load_libraries}
library(janitor)
library(phyloseq)  # BiocManager::install("phyloseq")
library(cowplot)
library(ggrepel)
library(scales)
library(RColorBrewer)
library(MASS)
library(lme4)
library(emmeans)
library(tidyverse)
library(vegan)
library(ggh4x)
library(stringi)
```

```{r ggplot_theme}
# Create custom ggplot theme
theme_custom <- function() {
  theme_bw(base_size = 10, base_family = "Arial") %+replace%
    theme(
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      panel.background = element_blank(),
      panel.border = element_rect(color = "black", fill = NA),
      legend.background = element_rect(fill = NA, colour = NA)
    )
}

sppord <- factor(levels = c("Cnat", "Dlab", "Pstr", "Pfur",
                            "Ssid", "Oann", "Ofav", "Ofra", "Mcav"))

## ggplot labeller
sppnames <- c(
  Cnat = "C. natans",
  Dlab = "D. labyrinthiformis",
  Pstr = "P. strigosa",
  Pfur = "P. furcata",
  Ssid = "S. siderea",
  Oann = "O. annularis",
  Ofav = "O. faveolata",
  Ofra = "O. franksi",
  Mcav = "M. cavernosa"
)

sitenames <- c(
  EastTektite = "ET",
  WhitePoint = "WP",
  CabritteHorn = "CH"
)

global_labeller <- labeller(
  gspp = sppnames,
  host_site = sitenames
)
```

# Load Data
```{r load_its2_profiles}
sam0 <- read_csv("Data/STJ_Metadata_2017_2019.csv") %>%
  clean_names()
sam1 <- as.matrix(sam0)
rownames(sam1) <- sam0$sample_name
sam <- sample_data(data.frame(sam1))

tax0 <- read_tsv(
  file = "Data/20200623_lenz/its2_type_profiles/111_20200624_2020-06-24_03-40-50.238182.profiles.absolute.abund_and_meta.txt",
  n_max = 6) %>%
  dplyr::select(-2) %>% 
  gather(UID, value, -1) %>% 
  spread(1, value) %>%
  clean_names()
tax1 <- as.matrix(tax0[, -1], dimnames = list(tax0$uid, colnames(tax0[-1])))
rownames(tax1) <- tax0$uid
tax <- tax_table(tax1)

otu0 <- read_tsv(
  file = "Data/20200623_lenz/its2_type_profiles/111_20200624_2020-06-24_03-40-50.238182.profiles.absolute.abund_and_meta.txt") %>% 
  rename(sample_name = "...2") %>%
  select(-1) %>%
  slice(7:n()) %>%
  mutate_at(2:ncol(.), as.numeric)
otu1 <- as.matrix(otu0[, -1])
rownames(otu1) <- otu0$sample_name
otu <- otu_table(otu1, taxa_are_rows = FALSE)

profiles <- phyloseq(otu, tax, sam)
```

```{r load_its2_variants}
taxnames <- read_tsv(
  file = "Data/20200623_lenz/post_med_seqs/111_20200624_2020-06-24_03-40-50.238182.seqs.absolute.abund_and_meta.txt",
  n_max = 0) %>%
  select(-(1:33)) %>%
  names(.)
tax0 <- data_frame(
  DIV = taxnames,
  clade = str_extract(DIV, "[A-Z]")
)
tax1 <- as.matrix(tax0)
rownames(tax1) <- tax0$DIV
tax <- tax_table(tax1)

otu0 <- read_tsv(
  file = "Data/20200623_lenz/post_med_seqs/111_20200624_2020-06-24_03-40-50.238182.seqs.absolute.abund_and_meta.txt") %>% 
  select(-1, -(3:33))
otu1 <- as.matrix(otu0[, -1])
rownames(otu1) <- otu0$sample_name
otu <- otu_table(otu1, taxa_are_rows = FALSE)

variants <- phyloseq(otu, tax, sam)
```

```{r its2_filter}
# Filter out RANDOM samples (non-tagged colonies)
variants <- subset_samples(variants, host_id != "RANDOM")
variants <- prune_taxa(taxa_sums(variants) != 0, variants)
profiles <- subset_samples(profiles, host_id != "RANDOM")
profiles <- prune_taxa(taxa_sums(profiles) != 0, profiles)
```

# Coral sampling summary
```{r sampling_summary_table1}
# Create table of number of tagged colonies sampled at each site at each year
table1 <- as_tibble(sample_data(profiles)) %>%
  drop_na(host_id) %>%
  group_by(year, host_species, host_site) %>%
  summarize(n = n()) %>%
  arrange(host_site, year) %>%
  pivot_wider(names_from = c(host_site, year), values_from = n) %>% 
  adorn_totals(where = "row")

# Save Table 1
write_csv(table1, path = "Output/Table1.csv")

# Display Table 1
knitr::kable(table1)
```

# ITS2 sequence dataset summary and QC
```{r its2_qc}
# Total number of post-MED sequences in dataset
sum(sample_sums(variants))
# Total number of samples in dataset
nsamples(variants)

# Filter out samples with low read counts
low <- sample_sums(variants) < 1000
table(low)  # Number of samples being filtered out due to low read counts
variants <- subset_samples(variants, !low)
variants <- prune_taxa(taxa_sums(variants) != 0, variants)
profiles <- subset_samples(profiles, !low)
profiles <- prune_taxa(taxa_sums(profiles) != 0, profiles)

# Read count analysis
hist(sample_sums(variants), breaks = 100)
# One sample is an outlier with very high read count -- remove for summary stats of remaining samples
outlier <- sample_sums(variants)[which(sample_sums(variants) == max(sample_sums(variants)))]
woout <- sample_sums(variants)[which(sample_sums(variants) != max(sample_sums(variants)))]

# Histogram of remaining samples read counts
hist(woout, breaks = 50)

# Mean number of sequences per sample and standard deviation
mean(woout); sd(woout)

# Count number of sequences and profiles by genus(/clade)
# Sequences
data.frame(tax_table(variants)) %>% as_tibble() %>%
  count(clade) %>%
  adorn_totals(where = "row")
# Profiles
data.frame(tax_table(profiles)) %>% as_tibble() %>%
  count(clade) %>%
  adorn_totals(where = "row")
```

# Analyze ITS2 sequence composition across all samples

### Barplots of sequence variant and profile composition
```{r its2_barplots, fig.width = 10, fig.height = 5}
## Transform to relative abundance
v <- transform_sample_counts(variants, function(x) x/sum(x))     # Relative abundance
p <- transform_sample_counts(profiles, function(x) x/sum(x))     # Relative abundance

## Filter out variants less than 0.1% relative abundance for plotting
vt <- transform_sample_counts(v, function(x) ifelse(x > 0.001, x, 0))
vt <- prune_taxa(taxa_sums(vt) != 0, vt)

## Order colonies by sequence variant composition similarity within species for plotting
mp <- merge_samples(vt, "host_id")
hc <- hclust(dist(data.frame(otu_table(mp))))
neword <- tibble(host_id = as.character(sample_data(mp)$host_id), ord = hc$order)

## Set colors for plotting sequence variants, with distinct palettes for each symbiont clade/genus
tt <- as_tibble(data.frame(tax_table(vt))) %>%
  arrange(clade, DIV)
nt <- tt %>% count(clade) %>% pivot_wider(names_from = clade, values_from = n)
acols <- colorRampPalette(brewer.pal(9, "Greys"))(nt$A)
bcols <- colorRampPalette(brewer.pal(9, "Greens"))(nt$B)
ccols <- colorRampPalette(brewer.pal(9, "Blues"))(nt$C)
dcols <- colorRampPalette(brewer.pal(4, "Oranges"))(nt$D)
set.seed(556)
vpal <- c(acols[shuffle(acols)], bcols[shuffle(bcols)], ccols[shuffle(ccols)], dcols[shuffle(dcols)])
vpal <- setNames(vpal, tt$DIV)

## Set plotting order for sequence variants
divord <- as_tibble(data.frame(tax_table(vt))) %>%
  arrange(clade, DIV) %>%
  mutate(DIV = as_factor(DIV))

vmdf <- psmelt(vt) %>%
  left_join(neword) %>%
  arrange(host_genus, host_species, host_site, ord, year, clade, DIV) %>%
  mutate(host_id = as_factor(host_id),
         sample_id = as_factor(paste(host_genus, host_species, host_id, year, sep = "_")),
         genus_species = as_factor(paste(str_sub(host_genus, 1, 1), ". ", host_species, sep = "")),
         gspp = factor(paste0(str_sub(host_genus, 1, 1), str_sub(host_species, 1, 3)),
                       levels = levels(sppord)),
         DIV = factor(DIV, levels = divord$DIV))

# Create barplot for sequence variants
g1 <- ggplot(vmdf, aes(x = year, y = Abundance)) + 
  geom_col(width = 1, aes(fill = DIV)) + 
  facet_nested(~ gspp + host_site + host_id, 
               labeller = global_labeller,
               space = "free_x",
               nest_line = element_line(linetype = 1),
               remove_labels = "x") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  ylab("ITS2 sequences") +
  scale_fill_manual(values = vpal[vmdf$OTU]) +
  theme(legend.position = "none", panel.background = element_blank(),
        text = element_text(size = 6), plot.margin = margin(5.5, 5.5, 1, 5.5),
        strip.text.x = element_text(margin = margin(0.5, 0, 0.5, 0, "mm"), face = "italic"),
        axis.title.x = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.spacing = unit(0.25, "mm"),
        strip.background = element_blank())

# Repeat for profiles
### truncate its2_type_profile names
pp <- as_tibble(data.frame(tax_table(p))) %>%
  arrange(clade, its2_type_profile) %>%
  mutate(its2_type_profile_trunc = str_trunc(its2_type_profile, 22, "right"))
np <- pp %>% count(clade) %>% pivot_wider(names_from = clade, values_from = n)
acols <- brewer.pal(np$A, "Greys")
bcols <- colorRampPalette(brewer.pal(9, "YlGn"))(np$B)
ccols <- colorRampPalette(brewer.pal(6, "PuBu"))(np$C)
dcols <- brewer.pal(np$D, "Reds")
set.seed(556)
ppal <- c(acols, bcols, ccols, dcols)
ppal <- setNames(ppal, pp$its2_type_profile_trunc)

proford <- as_tibble(data.frame(tax_table(p))) %>%
  mutate(its2_type_profile_trunc = str_trunc(its2_type_profile, 22, "right")) %>%
  arrange(clade, its2_type_profile_trunc) %>%
  mutate(its2_type_profile = as_factor(its2_type_profile_trunc))
  
pmdf <- psmelt(p) %>%
  left_join(neword) %>%
  arrange(host_genus, host_species, host_site, ord, year, clade, its2_type_profile) %>%
  mutate(host_id = as_factor(host_id),
         its2_type_profile_trunc = str_trunc(its2_type_profile, 22, "right"),
         sample_id = as_factor(paste(host_genus, host_species, host_id, year, sep = "_")),
         genus_species = as_factor(paste(str_sub(host_genus, 1, 1), ". ", host_species, sep = "")),
         gspp = factor(paste0(str_sub(host_genus, 1, 1), str_sub(host_species, 1, 3)),
                       levels = levels(sppord))) %>%
  filter(Abundance > 0.01)

# Create barplot for profiles
g2 <- ggplot(pmdf, aes(x = year, y = Abundance)) +
    geom_col(width = 1, aes(fill = its2_type_profile_trunc)) +
    facet_nested(~ gspp + host_site + host_id) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    ylab("Profiles") +
    scale_fill_manual(values = ppal[pmdf$its2_type_profile_trunc], limits = force, drop = TRUE) +
    theme(legend.position = "bottom", legend.key.size = unit(3, "mm"), legend.title = element_blank(),
          legend.margin = margin(1, 0, 0, 0), legend.box.margin = margin(-10, -10, 0, -10),
          legend.key = element_rect(fill = NA),
          plot.margin = margin(0, 5.5, 5.5, 5.5),
          text = element_text(size = 6), legend.text = element_text(size = 4),
          panel.background = element_blank(), strip.text.x = element_blank(),
          axis.title.x = element_blank(), axis.text.y = element_blank(), 
          axis.text.x = element_text(angle = 90, vjust = 0.5, size = 3),
          axis.ticks.x = element_line(size = 0.25),
          axis.ticks.y = element_blank(),
          strip.background = element_blank(),
          panel.spacing = unit(0.25, "mm")) +
    guides(fill = guide_legend(ncol = 3))

# Extract legend from profiles barplot to use in multipanel figure below
g2leg <- get_legend(g2)
g2leg <- ggplotify::as.ggplot(g2leg) + theme(plot.margin = unit(c(-30, 0, 0, -10), "mm"))
g2 <- g2 + theme(legend.position = "none")

# Combine variants and profiles barplots into single panel to be Figure 1a
fig1a <- plot_grid(g1, g2, nrow = 2, label_fontface = "plain", rel_heights = c(0.6, 0.4))

# Display Figure 1a
fig1a
```

### Network analysis of dominant symbiont profiles
```{r network_analysis}
library(igraph)
# only one connection per unique colony
df2 <- as_tibble(pmdf) %>%
  filter(Abundance > 0.5) %>%   # Dominant symbiont profiles only
  select(its2_type_profile_trunc, host_species, host_id, year) %>%
  distinct(its2_type_profile_trunc, host_species, host_id) %>%
  group_by(its2_type_profile_trunc, host_species) %>%
  mutate(n = n(),
         weights = 1/n) %>%
  ungroup()

net <- graph_from_data_frame(df2, directed = FALSE)

# Modify vertex attributes for plotting
V(net)$type <- if_else(V(net)$name %in% unique(pmdf$host_species), "coral", "sym")
V(net)$clade <- if_else(V(net)$type == "sym", str_sub(V(net)$name, 1, 1), NA_character_)
V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
ppal2 <- c(acols, bcols, ccols, dcols)
ppal2 <- setNames(ppal2, pp$its2_type_profile_trunc)
V(net)$color <- ifelse(is.na(V(net)$clade), NA, ppal2[V(net)$name])
V(net)$label <- sapply(V(net)$name, function(x) paste0(stri_wrap(str = x, width = 10, whitespace_only = FALSE),
          collapse = "\n"))
V(net)$label[29:37] <- c("C. natans", "D. labyrinthiformis", "M. cavernosa", "O. annularis", 
                         "O. faveolata", "O. franksi", "P. furcata", "P. strigosa", "S. siderea")
V(net)$label.font <- if_else(V(net)$type == "coral", 3, 1)
V(net)$label.cex = if_else(V(net)$type == "coral", 0.5, 0.3)
V(net)$label.color = "black"
V(net)$frame.color = NA
V(net)$size = if_else(V(net)$type == "coral", 8, 12)

# Plot network
library(ggplotify)
set.seed(1)
mynet <- base2grob(~plot(net, layout = layout_with_fr(net, weights = E(net)$weights^1), edge.width = 0.5))
ggnet <- ggplotify::as.ggplot(mynet) +
  theme(plot.margin = unit(c(-20, -40, 60, -100), "points"))

# Display network
ggnet
```

### Create and save multipanel Figure 1 with barplots and network plot
```{r Figure1}
fig1a <- fig1a + theme(plot.margin = unit(c(5, 0, 0, 0), "mm"))
fig1b <- plot_grid(g2leg, ggnet, labels = c("", "b)"), label_x = -0.15) + 
  theme(plot.background = element_rect(fill = "white", color = NA))
fig1 <- plot_grid(fig1a, fig1b, nrow = 2, rel_heights = c(0.4, 0.6), labels = c("a)", ""), label_y = 0.9, label_x = -0.007) + 
  theme(plot.margin = unit(c(0, 0, -27, 0), "mm"))

ggsave(fig1, filename = "Output/Figure1.png", width = 183, height = 120, units = "mm", dpi = 600)
```

# Analyze coral host-specificity of sequence variants within each symbiont genus

### Visualize sequence assemblage variation using NMDS
```{r nmds_Figure2, fig.width = 10, fig.height = 3}
# Get list of samples with >5% relative abundance of each symbiont genus
samples_clades <- vmdf %>%
  # Get clade level relative abundances
  group_by(Sample, gspp, host_site, host_id, year, clade) %>%
  summarize(Abundance = sum(Abundance)) %>%
  group_by(Sample, gspp, host_site, host_id, year) %>%
  ungroup()
samples_with_5pA <- samples_clades %>% filter(clade == "A", Abundance >= 0.05)
samples_with_5pB <- samples_clades %>% filter(clade == "B", Abundance >= 0.05)
samples_with_5pC <- samples_clades %>% filter(clade == "C", Abundance >= 0.05)
samples_with_5pD <- samples_clades %>% filter(clade == "D", Abundance >= 0.05)

# Ordination analysis of variants within each clade/genus for samples with that clade at >5% relative abundance

# Symbiodinium
varA <- subset_samples(variants, sample_name %in% samples_with_5pA$Sample)
varA <- subset_taxa(varA, clade == "A")
# Run NMDS for visualization
set.seed(54)
varA.nmds <- ordinate(varA, "NMDS", "bray", trace = FALSE)
varA.nmds.s <- plot_ordination(varA, varA.nmds, type="samples")

# Breviolum
varB <- subset_samples(variants, sample_name %in% samples_with_5pB$Sample)
varB <- subset_taxa(varB, clade == "B")
# Run NMDS for visualization
set.seed(54)
varB.nmds <- ordinate(varB, "NMDS", "bray", trace = FALSE)
varB.nmds.s <- plot_ordination(varB, varB.nmds, type="samples")

# Cladocopium
varC <- subset_samples(variants, sample_name %in% samples_with_5pC$Sample)
varC <- subset_taxa(varC, clade == "C")
# Run NMDS for visualization
set.seed(55)
varC.nmds <- ordinate(varC, "NMDS", "bray", trace = FALSE)
varC.nmds.s <- plot_ordination(varC, varC.nmds, type="samples")

# Durusdinium
varD <- subset_samples(variants, sample_name %in% samples_with_5pD$Sample)
varD <- subset_taxa(varD, clade == "D")
# Run NMDS for visualization
set.seed(66)
varD.nmds <- ordinate(varD, "NMDS", "bray", trace = FALSE)
varD.nmds.s <- plot_ordination(varD, varD.nmds, type="samples")

# Plot NMDS for each genus
clade_nmds <- bind_rows(
  Symbiodinium = varA.nmds.s$data,
  Breviolum = varB.nmds.s$data, 
  Cladocopium = varC.nmds.s$data, 
  Durusdinium = varD.nmds.s$data,
  .id = "sym_genus"
) %>%
  arrange(host_id, year) %>%
  mutate(sym_genus = factor(sym_genus, levels = c("Symbiodinium", "Breviolum", 
                                                  "Cladocopium", "Durusdinium"))) %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = host_species, shape = host_species)) +
  facet_wrap(~ sym_genus, scales = "free", nrow = 1) +
  geom_point(alpha = 1, size = 2) +
  geom_path(aes(group = host_id), linewidth = 0.2) +
  scale_shape_manual(values = 8:0, labels = c("O. annularis", "M. cavernosa", "O. faveolata", "O. franksi", "P. furcata",
                                              "D. labyrinthiformis", "C. natans", "S. siderea", "P. strigosa")) +
  scale_color_discrete(labels = c("O. annularis", "M. cavernosa", "O. faveolata", "O. franksi", "P. furcata",
                                              "D. labyrinthiformis", "C. natans", "S. siderea", "P. strigosa")) +
  theme_custom() +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        strip.text = element_text(face = "italic"),
        legend.key.size = unit(2, "mm"), legend.title = element_blank(),
          legend.margin = margin(1, 0, 0, 0), legend.box.margin = margin(-10, -10, 0, -10),
          legend.key = element_rect(fill = NA),
          text = element_text(size = 6), legend.text = element_text(size = 4, face = "italic")) +
  guides(shape = guide_legend(ncol = 9))

# Save Figure 2
ggsave(clade_nmds, filename = "Output/Figure2.png", width = 235, height = 77, units = "mm")

# Display Figure 2
clade_nmds
```

### Test for differences between host species using pairwise PERMANOVAs
```{r}
# Pairwise PERMANOVAs
devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

# Test whether Symbiodinium sequence assemblages differ among host species hosting them at >5%
Amod <- pairwiseAdonis::pairwise.adonis(distance(varA, "bray"), 
                                        as_tibble(sample_data(varA))$host_species,
                                        p.adjust.m = "bonferroni")
knitr::kable(Amod)
# No significant differences for Symbiodinium (low sample sizes)



# Test whether Breviolum sequence assemblages differ among host species hosting them at >5%
Bmod <- pairwiseAdonis::pairwise.adonis(distance(varB, "bray"), 
                                        as_tibble(sample_data(varB))$host_species,
                                        p.adjust.m = "fdr")
knitr::kable(Bmod)

# Save statistical results as Table S1
write_excel_csv(Bmod, path = "output/TableS1.csv")

# Visualize pairwise test significance as heatmap
Bmod <- Bmod %>%
  separate(col = pairs, into = c("var1", "var2"), sep = " vs ")
figS1 <- Bmod %>%
  bind_rows(Bmod %>% rename(var1 = var2, var2 = var1)) %>%
  full_join(expand_grid(var1 = unique(as_tibble(sample_data(varB))$host_species), 
                        var2 = unique(as_tibble(sample_data(varB))$host_species))) %>%
  mutate(p.sig = case_when(p.adjusted < 0.05 ~ p.adjusted, TRUE ~ NA_real_)) %>%
  ggplot(aes(x = var1, y = var2)) +
  geom_tile(col = "black", aes(fill = log10(p.adjusted))) +
  geom_text(aes(label = sig), fontface = "bold", size = 3) +
  labs(x = "", y = "") +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(2, "mm"),
        text = element_text(size = 4), legend.text = element_text(size = 4),
        legend.title = element_text(size = 4)) +
  scale_fill_gradient(low = "red", high = "white"); figS1

# Save heatmap as Figure S2
ggsave(figS1, filename = "output/FigureS2.png", device = "png", width = 50, height = 30, units = "mm")



# Test whether Cladocopium sequence assemblages differ among host species hosting them at >5%
Cmod <- pairwiseAdonis::pairwise.adonis(distance(varC, "bray"), 
                                        as_tibble(sample_data(varC))$host_species,
                                        p.adjust.m = "fdr")
knitr::kable(Cmod)

# Save statistical results as Table S2
write_excel_csv(Cmod, path = "output/TableS2.csv")

# Visualize pairwise test significance as heatmap
Cmod <- Cmod %>%
  separate(col = pairs, into = c("var1", "var2"), sep = " vs ")
figS2 <- Cmod %>%
  bind_rows(Cmod %>% rename(var1 = var2, var2 = var1)) %>%
  full_join(expand_grid(var1 = unique(as_tibble(sample_data(varC))$host_species), 
                        var2 = unique(as_tibble(sample_data(varC))$host_species))) %>%
  mutate(p.sig = case_when(p.adjusted < 0.05 ~ p.adjusted, TRUE ~ NA_real_)) %>%
  ggplot(aes(x = var1, y = var2)) +
  geom_tile(col = "black", aes(fill = log10(p.adjusted))) +
  geom_text(aes(label = sig), fontface = "bold", size = 3) +
  labs(x = "", y = "") +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(2, "mm"),
        text = element_text(size = 4), legend.text = element_text(size = 4),
        legend.title = element_text(size = 4)) +
  scale_fill_gradient(low = "red", high = "white"); figS2

# Save heatmap as Figure S3
ggsave(figS2, filename = "output/FigureS3.png", device = "png", width = 60, height = 40, units = "mm")


# Test whether Durusdinium sequence assemblages differ among host species hosting them at >5%
Dmod <- pairwiseAdonis::pairwise.adonis(distance(varD, "bray"), 
                                        as_tibble(sample_data(varD))$host_species,
                                        p.adjust.m = "fdr")
knitr::kable(Dmod)
# No significant differences for Durusdinium (low sample sizes)
```

# Analyze symbiont community change over time within colonies

### Find number of corals that changed dominant genus or profile
```{r, change_dominant}
# Find corals that changed dominant genus
genus_change <- pmdf %>%
  filter(Abundance >= 0.5) %>%
  group_by(gspp, host_id) %>%
  summarize(n_dom_clades = n_distinct(host_id, clade)) %>%
  filter(n_dom_clades > 1)
pmdf %>% 
  filter(Abundance >= 0.5, host_id %in% genus_change$host_id) %>%
  select(host_id, host_species, year, clade, its2_type_profile)

# Find corals that changed dominant profile within a genus
prof_change <- pmdf %>%
  filter(Abundance >= 0.5) %>%
  group_by(gspp, clade, host_id) %>%
  summarize(n_dom_profs = n_distinct(host_id, clade, its2_type_profile)) %>%
  filter(n_dom_profs > 1) %>%
  ungroup()
n_distinct(prof_change$host_id)
# 13 colonies changed dominant profile within genus

# How many total colonies were sampled in more than one year?
pmdf %>%
  distinct(host_id, year) %>%
  count(host_id) %>%
  count(n > 1)
# 56 colonies sampled more than once, 4 shuffled genera, and 13 shuffled profiles, so 39 did not
```

### Find number of corals that hosted multiple symbiont genera
```{r mult_genera}
# Get abundance of each clade in each sample
clades <- vmdf %>%
  # Get clade level relative abundances
  group_by(gspp, host_site, host_id, year, clade) %>%
  summarize(Abundance = sum(Abundance)) %>%
  group_by(gspp, host_site, host_id, year) %>%
  pivot_wider(names_from = clade, values_from = Abundance) %>%
  ungroup()

# Get number of clades over threshold abundance in each sample
nclades <- clades %>%
  nest(data = c(A, B, C, D)) %>%
  mutate(nclades.0.1 = map_dbl(data, ~ sum(. > 0.001)),
         nclades.1   = map_dbl(data, ~ sum(. > 0.01)),
         nclades.10  = map_dbl(data, ~ sum(. > 0.1)))

# Find coral colonies that had multiple clades/genera (any time point)
mult <- nclades %>%
  group_by(gspp, host_site, host_id) %>%
  summarize(mult.0.1 = any(nclades.0.1 > 1),
            mult.1   = any(nclades.1 > 1),
            mult.10  = any(nclades.10 > 1))

table(mult$mult.0.1) 
# 64/70 (91%) colonies had multiple clades > 0.1% abund

table(mult$mult.1) 
# 33/70 (47%) colonies had multiple clades > 1% abund

table(mult$mult.10) 
# 11/70 (16%) colonies had multiple clades > 10% abund


# Find coral colonies that changed in the number of clades > 1%
nc1 <- nclades %>%
  group_by(host_id) %>%
  filter(n() > 1) %>%
  summarize(nc1 = n_distinct(nclades.1))

table(nc1$nc1) 
# 27/56 corals sampled more than once showed a change in the number of clades >1% relative abundance
```

### Visualize symbiont community change over time using NMDS
```{r change_nmds_Figure3, fig.width = 6, fig.height = 6}
### Visualize change within colonies over time
varr <- transform_sample_counts(variants, function(x) sqrt(x/sum(x))) 
#set.seed(55)
set.seed(56)
varr.nmds <- ordinate(varr, "NMDS", "bray", trace = FALSE)
varr.nmds.s <- plot_ordination(varr, varr.nmds, type="samples")

nmds_fig <- as_tibble(varr.nmds.s$data) %>%
  mutate(gspp = factor(paste0(str_sub(host_genus, 1, 1), str_sub(host_species, 1, 3)))) %>%
  arrange(host_id, year) %>%
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(shape = year, color = year), alpha = 0.6) +
  geom_path(aes(group = host_id), alpha = 0.5, arrow = arrow(ends = "last", length = unit(1, "mm")), lwd = 0.25) +
  facet_wrap(~gspp, labeller = global_labeller) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb')) +
  theme_custom() +
  theme(strip.background = element_blank(),
          strip.text = element_text(face = "italic"),
          text = element_text(size = 7), legend.text = element_text(size = 4),
          axis.text.x = element_text(hjust = 0)) +
  theme(legend.position = "none")

# Save Figure 3
ggsave(nmds_fig, filename = "Output/Figure3.png", width = 85, height = 85, units = "mm")

# Display Figure 3
nmds_fig
```

### Analyze the change in *Breviolum* in *M. cavernosa*
```{r}
# Mcav Breviolum change over time
vmdf %>%
  filter(host_species == "cavernosa", clade == "B") %>%
  group_by(host_id, year) %>%
  summarize(totB = sum(Abundance)) %>%
  group_by(year) %>%
  summarize(`mean % Breviolum` = mean(totB)) %>%
  knitr::kable()
```

# Compare between-colony and within-colony Bray-Curtis dissimilarities across species
```{r between_within}
varr <- transform_sample_counts(variants, function(x) sqrt(x/sum(x)))    # Square-root transform

bd <- as.matrix(distance(varr, "bray"))
xy <- t(combn(colnames(bd), 2))
bd2 <- as_tibble(data.frame(xy, dist=bd[xy])) %>%
  separate(X1, into = c("site1", "species1", "host_id1", "year1")) %>%
  separate(X2, into = c("site2", "species2", "host_id2", "year2")) %>%
  filter(species1 == species2) %>%    # only interested in within-species differences
  mutate(group = case_when(host_id1 == host_id2 ~ "within", TRUE ~ "between")) %>%
  filter(group == "within" |
         group == "between" & year1 == year2)



#order species by highest median between-colony difference
brayord <- bd2 %>%
  filter(group == "between") %>%
  group_by(species1) %>%
  summarize(medbray = median(dist)) %>%
  arrange(-medbray) %>%
  mutate(gspp = as_factor(species1))
  
# Make raincloud plot
source("https://raw.githubusercontent.com/datavizpyr/data/master/half_flat_violinplot.R")
f4 <- bd2 %>% mutate(gspp = factor(species1, levels = levels(brayord$gspp))) %>%
  ggplot(aes(x = group, y = dist)) +
    facet_wrap(~ gspp, nrow = 2, labeller = global_labeller) +
    geom_flat_violin(adjust = 1, position = position_nudge(x = 0.15), scale = "width",
                     color = NA, fill = "grey72") +
    geom_boxplot(width = 0.1, outlier.shape = NA, lwd = 0.2,
                 position = position_nudge(x = 0.15)) +
    geom_jitter(width = 0.075, alpha = 0.2, stroke = 0) +
    theme_custom() +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "italic"),
          text = element_text(size = 7), legend.text = element_text(size = 4),
          axis.text.x = element_text(hjust = 0)) +
    labs(y = "Bray-Curtis Dissimilarity", x = "") +
    coord_cartesian(xlim = c(1.3, 2.1))

# Save Figure 4
ggsave(f4, filename = "Output/Figure4.png", width = 112, height = 77, units = "mm")

# Display Figure 4
f4

# Compare magnitude of between- to within-colony variation
bd2 %>%
  group_by(species1, group) %>%
  summarize(median_dist = median(dist)) %>%
  pivot_wider(names_from = group, values_from = median_dist) %>%
  mutate(btw_wit = between/within) %>%
  arrange(btw_wit)
```

