---
title: "St. John symbiont time series"
author: "R. Cunning"
date: "2/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(janitor)
library(phyloseq)  # BiocManager::install("phyloseq")
library(cowplot)
library(ggrepel)
library(scales)
library(RColorBrewer)
```


```{r profiles, include = FALSE}
sam0 <- read_csv("Data/STJ_Metadata_2017_2019.csv") %>%
  clean_names()
sam1 <- as.matrix(sam0)
rownames(sam1) <- sam0$sample_name
sam <- sample_data(data.frame(sam1))

tax0 <- read_tsv(
  file = "Data/20200623_lenz/its2_type_profiles/111_20200624_2020-06-24_03-40-50.238182.profiles.absolute.abund_and_meta.txt",
  n_max = 6) %>%
  dplyr::select(-2) %>% 
  gather(UID, value, -1) %>% 
  spread(1, value) %>%
  clean_names()
tax1 <- as.matrix(tax0[, -1], dimnames = list(tax0$uid, colnames(tax0[-1])))
rownames(tax1) <- tax0$uid
tax <- tax_table(tax1)

otu0 <- read_tsv(
  file = "Data/20200623_lenz/its2_type_profiles/111_20200624_2020-06-24_03-40-50.238182.profiles.absolute.abund_and_meta.txt") %>% 
  rename(sample_name = X2) %>%
  select(-1) %>%
  slice(7:n()) %>%
  mutate_at(2:ncol(.), as.numeric)
otu1 <- as.matrix(otu0[, -1])
rownames(otu1) <- otu0$sample_name
otu <- otu_table(otu1, taxa_are_rows = FALSE)

profiles <- phyloseq(otu, tax, sam)
```

```{r variants, include = FALSE}
taxnames <- read_tsv(
  file = "Data/20200623_lenz/post_med_seqs/111_20200624_2020-06-24_03-40-50.238182.seqs.absolute.abund_and_meta.txt",
  n_max = 0) %>%
  select(-(1:33)) %>%
  names(.)
tax0 <- data_frame(
  DIV = taxnames,
  clade = str_extract(DIV, "[A-Z]")
)
tax1 <- as.matrix(tax0)
rownames(tax1) <- tax0$DIV
tax <- tax_table(tax1)

otu0 <- read_tsv(
  file = "Data/20200623_lenz/post_med_seqs/111_20200624_2020-06-24_03-40-50.238182.seqs.absolute.abund_and_meta.txt") %>% 
  select(-1, -(3:33))
otu1 <- as.matrix(otu0[, -1])
rownames(otu1) <- otu0$sample_name
otu <- otu_table(otu1, taxa_are_rows = FALSE)

variants <- phyloseq(otu, tax, sam)
```

```{r save, include = FALSE}
save(profiles, variants, file = "data/Symb_phyloseq.RData")
```

```{r QC, include = FALSE}
ntaxa(variants)
ntaxa(profiles)

min(sample_sums(variants))
max(sample_sums(variants))
hist(log10(sample_sums(variants)))
mean(sample_sums(variants))
sd(sample_sums(variants))/sqrt(nsamples(variants))

# Filter out samples with low read counts
low <- sample_sums(variants) < 1000
variants <- subset_samples(variants, !low)
variants <- prune_taxa(taxa_sums(variants) != 0, variants)
profiles <- subset_samples(profiles, !low)
profiles <- prune_taxa(taxa_sums(profiles) != 0, profiles)

ntaxa(variants)
ntaxa(profiles)



# Transform to relative abundance
#variants <- transform_sample_counts(variants, fun = function(x) x / sum(x))
#profiles <- transform_sample_counts(profiles, fun = function(x) x / sum(x))
```

```{r color_palettes, include = FALSE}
# # Subset colonies sampled in all three years
# all3 <- sample_data(profiles) %>%
#   group_by(host_id) %>%
#   filter(n() == 3) %>%
#   distinct(host_id) %>%
#   pull(host_id)

# Create standardized color pallettes for plotting profiles and variants

#ppal <- sample(allcolors, ntaxa(profiles))
#ppal <- hue_pal(l = 65)(ntaxa(profiles))
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
set.seed(555)
ppal <- sample(col_vector, ntaxa(profiles))
ppal <- setNames(ppal, data.frame(tax_table(profiles))$its2_type_profile)

allcolors <- grDevices::colors()
set.seed(888)
vpal <- sample(allcolors, ntaxa(variants))
#vpal <- hue_pal(l = 65)(ntaxa(variants))
vpal <- setNames(vpal, taxa_names(variants))
```

```{r plotting_functions, include = FALSE}
# Plotting functions
plot_both <- function(mycol) {
  s <- subset(sample_data(variants), host_id == mycol)
  v <- merge_phyloseq(tax_table(variants), otu_table(variants), s)
  p <- merge_phyloseq(tax_table(profiles), otu_table(profiles), s)
  v <- prune_taxa(taxa_sums(v) != 0, v)
  p <- prune_taxa(taxa_sums(p) != 0, p)
  vmdf <- psmelt(v)
  seqdepth <- vmdf %>%
    group_by(year) %>%
    summarize(sum = paste0("n=", sum(Abundance)))
  ti <- paste0(vmdf[1, c("host_genus", "host_species", "host_id", "host_site")], collapse = " ")
  v_plot <- ggplot(vmdf, aes(x = year, y = Abundance, fill = DIV, label = DIV)) + 
    geom_bar(stat = "identity", position = "fill", color = "black", lwd = 0.1) + 
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0, .1)), labels = scales::percent) +
    scale_fill_manual(values = vpal[vmdf$OTU]) +
    geom_text(aes(size = Abundance), position = position_fill(vjust = 0.5)) +
    geom_text(data = seqdepth, aes(x = year, y = 1.01, label = sum), vjust = 0, size = 3, inherit.aes = FALSE) +
    theme(legend.position = "none", panel.background = element_blank(), 
          axis.title.x = element_blank(), axis.title.y = element_blank()) +
    coord_cartesian(ylim = c(0, 1))
  pmdf <- psmelt(p)
  p_plot <- ggplot(pmdf, aes(x = year, y = Abundance, fill = its2_type_profile, label = its2_type_profile)) +
    geom_bar(stat = "identity", position = "fill", color = "black", lwd = 0.1) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0, .1)), labels = scales::percent) +
    scale_fill_manual(values = ppal[pmdf$its2_type_profile]) +
    geom_label_repel(position = position_fill(vjust = 0.5), force = 2,
                     point.size = NA, xlim = c(-0.1, 3.6), ylim = c(0.1,0.9), size = 2) +
    theme(legend.position = "none", panel.background = element_blank(), 
          axis.title.x = element_blank(), axis.title.y = element_blank()) +
    coord_cartesian(clip = "off")
  plot_row <- plot_grid(p_plot, v_plot, labels = c("ITS2 Profiles", "ITS2 Variants"), label_fontface = "plain")
  title <- ggdraw() + 
    draw_label(ti, fontface = 'bold', x = 0, hjust = 0) +
    theme(plot.margin = margin(0, 0, 0, 7))
  plot_grid(title, plot_row, ncol = 1, rel_heights = c(0.1, 1))
}

#plot_both(mycol = all3[1])

# Plot by species and site
# ordf <- data.frame(sam) %>%
#   distinct(host_id, host_genus, host_species, host_site) %>%
#   filter(host_id != "RANDOM") %>%
#   filter(host_id %in% all3) %>%
#   arrange(host_genus, host_species, host_site)



# new plotting function to condense by species rather than by individual
plot_both2 <- function(...) {
  s <- subset(sample_data(variants), ...)
  v <- merge_phyloseq(tax_table(variants), otu_table(variants), s)
  p <- merge_phyloseq(tax_table(profiles), otu_table(profiles), s)
  v <- prune_taxa(taxa_sums(v) != 0, v)
  p <- prune_taxa(taxa_sums(p) != 0, p)
  vmdf <- psmelt(v)
  seqdepth <- vmdf %>%
    group_by(year) %>%
    summarize(sum = paste0("n=", sum(Abundance)))
  ti <- paste0(vmdf[1, c("host_genus", "host_species")], collapse = " ")
  v_plot <- ggplot(vmdf, aes(x = year, y = Abundance, fill = DIV, label = DIV)) + 
    geom_bar(stat = "identity", position = "fill") + 
    facet_wrap(~ host_site + host_id, nrow = 1) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0, .1)), labels = scales::percent) +
    ylab("ITS2 sequences") +
    scale_fill_manual(values = vpal[vmdf$OTU]) +
    # geom_text(aes(size = Abundance), position = position_fill(vjust = 0.5)) +
    # geom_text(data = seqdepth, aes(x = year, y = 1.01, label = sum), vjust = 0, size = 3, inherit.aes = FALSE) +
    theme(legend.position = "none", panel.background = element_blank(), 
          axis.title.x = element_blank(), axis.text.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    coord_cartesian(ylim = c(0, 1))
  pmdf <- psmelt(p)
  p_plot <- ggplot(pmdf, aes(x = year, y = Abundance, fill = its2_type_profile, label = its2_type_profile)) +
    geom_bar(stat = "identity", position = "fill") +
    facet_wrap(~ host_site + host_id, nrow = 1) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0, .1)), labels = scales::percent) +
    ylab("Profiles") +
    scale_fill_manual(values = ppal[pmdf$its2_type_profile]) +
    # geom_label_repel(position = position_fill(vjust = 0.5), force = 2,
    #                  point.size = NA, xlim = c(-0.1, 3.6), ylim = c(0.1,0.9), size = 2) +
    theme(legend.position = "bottom", panel.background = element_blank(), 
          axis.title.x = element_blank(), axis.text.x = element_text(angle = 90),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          strip.background = element_blank(), strip.text.x = element_blank())# +
    #guides(fill = guide_legend(nrow = 4)) # +
    #coord_cartesian(clip = "off")
  plot_grid(v_plot, p_plot, nrow = 2, label_fontface = "plain")
}
```

# Summary figures for each species

The top row of bars shows all ITS2 sequence variants, while the bottom row of bars shows the associated ITS2 profiles. Legend refers to profiles only, not variants.

### *Colpophyllia natans*
```{r natans_summary, fig.width = 11}
plot_both2(host_species == "natans")
```

*C. natans* colonies at Cabritte Horn and East Tektite were dominated by a few different *Breviolum* profiles, with low levels of *Cladocopium* and *Durusdinium* that fluctuated randomly over time. At White Point, one colony was dominated by *Cladocopium* and one by *Durusdinium*.

### *Diploria labyrinithiformis*
```{r labrynthiformis_summary, fig.width = 11}
plot_both2(host_species == "labrynthiformis")
```

One colony at East Tektite (57) was dominated by *Cladocopium*, but otherwise *D. labyrinthiformis* was dominated by several distinct *Breviolum* profiles. These mostly remained the same over time within colonies, but changed for a few (colony 36, 37, 62). Will have to dig into these to see if these may be real changes or analytical artifacts. 

### *Montastraea cavernosa*
```{r cavernosa_summary, fig.width = 11}
plot_both2(host_species == "cavernosa")
```

*M. cavernosa* is dominated by a couple of different *Cladocopium* profiles with some background *Breviolum*. Two colonies at White Point have a distinct *Cladocopium* profile, but there doesn't appear to be any particular change over time.

### *Orbicella spp.*
```{r orbicella_summary, fig.width = 11}
plot_both2(host_genus == "Orbicella")
```

*Orbicella* colonies are sometimes dominated by *Symbiodinium*, *Breviolum*, and *Cladocopium*. Most stay the same over time, but some colonies (24 and 64, both at White Point), have mixtures of *Symbiodinium* and  *Cladocopium* in 2017, and then only *Cladocopium* in 2019. Another colony at White Point (17) appears to change between two *Cladocopium* profiles. At other sites, little evidence of change over time.

### *Pseudodiploria strigosa*
```{r strigosa_summary, fig.width = 11}
plot_both2(host_species == "strigosa")
```

*P. strigosa* colonies contain mostly *Breviolum* profiles, with some evidences for changes among *Breviolum* types over time. Two colonies at East Tektite (45 and 54) are dominated by *Cladocopium*.

### *Siderastrea siderea*
```{r siderea_summary, fig.width = 11}
plot_both2(host_species == "siderea")
```

*S. siderea* hosts mainly *Cladocopium* profiles that do not change over time. A couple of colonies (7 and 73) may have fluctuating levels of multiple *Cladocopium* types. One colony at Cabritte Horn (51) is dominated by *Breviolum*.

# Detailed figures for each coral colony

### Colpophyllia natans
```{r}
cnat <- data.frame(sam) %>%
  distinct(host_id, host_genus, host_species, host_site) %>%
  filter(host_species == "natans")

for (i in 1:nrow(cnat)) {
  p <- plot_both(mycol = cnat[i, "host_id"])
  print(p)
}
```

### Diploria labyrinthiformis
```{r}
dlab <- data.frame(sam) %>%
  distinct(host_id, host_genus, host_species, host_site) %>%
  filter(host_species == "labrynthiformis")

for (i in 1:nrow(dlab)) {
  p <- plot_both(mycol = dlab[i, "host_id"])
  print(p)
}
```

### Montastraea cavernosa
```{r}
mcav <- data.frame(sam) %>%
  distinct(host_id, host_genus, host_species, host_site) %>%
  filter(host_species == "cavernosa")

for (i in 1:nrow(mcav)) {
  p <- plot_both(mycol = mcav[i, "host_id"])
  print(p)
}
```

### Orbicella spp.
```{r}
orb <- data.frame(sam) %>%
  distinct(host_id, host_genus, host_species, host_site) %>%
  filter(host_genus == "Orbicella")

for (i in 1:nrow(orb)) {
  p <- plot_both(mycol = orb[i, "host_id"])
  print(p)
}
```

### Pseudodiploria strigosa
```{r}
pstr <- data.frame(sam) %>%
  distinct(host_id, host_genus, host_species, host_site) %>%
  filter(host_species == "strigosa")

for (i in 1:nrow(pstr)) {
  p <- plot_both(mycol = pstr[i, "host_id"])
  print(p)
}
```

### Siderastrea siderea
```{r}
ssid <- data.frame(sam) %>%
  distinct(host_id, host_genus, host_species, host_site) %>%
  filter(host_species == "siderea")

for (i in 1:nrow(ssid)) {
  p <- plot_both(mycol = ssid[i, "host_id"])
  print(p)
}
```





